---
title: "Microarray Analysis"
author: "Andrés Aravena"
date: "5 May 2015"
output:
  ioslides_presentation:
    keep_md: yes
    logo: ~/Src/blog/assets/IU_Logo_128.png
    self_contained: no
    smart: no
---

## Input data
We will use a modified version of the file that **Walid Khalilia** provided.

We read the data frame with this command
```{r read.csv, cache=TRUE}
m <- read.csv("microarray.csv")
```
The resulting data frame has `r nrow(m)` rows and `r ncol(m)` columns.

## Column names
```{r colnames}
colnames(m)
```

## Conditions to contrast
We have four conditions
```{r matrix, echo=FALSE, results='asis'}
mat <- sapply(strsplit(grep('AVG_Signal$', colnames(m), value = TRUE),".", fixed=TRUE), `[`, 1)
mat <- matrix(mat, 2, 2, dimnames=list(c(1,2),c("Treat", "Control")))
knitr::kable(as.data.frame(mat))
```

## Detecting change
If 

+ the signal intensity of treatment is $T$
+ the signal intensity of control $C$

Then the natural way to compare two conditions is to evaluate the difference $D$ as:
$$D=T-C$$

which in R is done like this
```{r difference}
d1 <- m$T1.AVG_Signal   - m$T1.AVG_Signal
```

## Percentual change
But this difference value depends too much of the scale.
When $D=100$ it may be for example

+ $T_1=50000$, $C_1=49900$
+ $T_2=200$, $C_2=100$

If we consider the percentage of change, we have
```{r pct}
T.1 <- 50000; C.1 <- 49900; (T.1-C.1)/C.1*100
T.2 <- 200; C.2 <- 100; (T.2-C.2)/C.2*100
```

## Fold Change
In the first case the change is only `r round(10000/C.1,1)`%.  
In the second case it is `r round(10000/C.2,1)`%.

So the difference is not a good way to compare two signals.

It is better to consider a ratio, the *fold change*
$$FC = \frac{T}{C}$$

+ if $T$ is twice of $C$, then $FC=2$
+ if $T$ is half of $C$, then $FC=0.5$
+ if $T$ is the same as $C$, then $FC=1$

## Fold Change is Asymetrical
```{r plot.fc}
T <- 1000; C <- c(250,500,1000, 2000, 4000, 8000, 16000)
plot(C, T/C, main="Fold Change")
```

## Log Fold Change
The solution to all this problems is to use *logarithms*.
```{r plot.lfc}
plot(C, log2(T/C), main="Log Fold Change")
```

## Alternative form
Notice that the previous formula can be written as
```{r plot.lfc2}
plot(C, log2(T) - log2(C), main="Log Fold Change")
```

## Applying to microarray data
Let us calculate *log fold change* for each one of the 3 replicas
```{r real.lfc, fig.height=3.5, cache=TRUE}
lfc.1 <- log2(m$T1.AVG_Signal)   - log2(m$C1.AVG_Signal)
lfc.2 <- log2(m$T1.AVG_Signal.1) - log2(m$C1.AVG_Signal.1)
lfc.3 <- log2(m$T1.AVG_Signal.2) - log2(m$C1.AVG_Signal.2)
hist(lfc.1, nclass=100)
```

## Average LFC
The vectors `lfc.1`, `lfc.2` and `lfc.3` contain three samples for each gene.

To average them we can use
```{r lfc0, eval=FALSE}
lfc <- rep(NA, length(lfc.1))
for(i in 1:length(lfc.1)) {
  lfc[i] <- mean(lfc.1[i], lfc.2[i], lfc.3[i])
}
```
but is more efficient to use the `mapply` function
```{r lfc1, cache=TRUE}
lfc <- mapply(mean, lfc.1, lfc.2, lfc.3)
```

## Hypothesis test
For each gene $i$, the *null hypothesis* is
$$LFC_i = 0 + \varepsilon_i$$
where each $\varepsilon_i$ represents the noise component following a normal distribution $N(0,\sigma^2)$.
We assume that the noise components of each gene are independent.

If the values for a gene are `x`, `y`, `z`, we test this hypothesis with this function
```{r def.p.val}
get.p.value <- function(x,y,z) {
  return( t.test(c(x,y,z), alternative = "two.sided")$p.value )
}
```

## *p*-values
If the null hypothesis is true, the probability of observing the log fold change is then
```{r calc.p.value, cache=TRUE}
p.val <- mapply(get.p.value, lfc.1, lfc.2, lfc.3)
```
(we used `mapply` again)
```{r hist.p.val, fig.height=3.5, echo=FALSE}
hist(p.val, nclass=20)
```

## Volcano Plot
Now we can compare the statistical and biological approaches
```{r volcano, cache=TRUE}
plot(lfc, -log10(p.val), pch=16, cex=0.4)
```
Lower *p*-values suggest that the null hypothesis is false

## Multiple test
Let us reject the null hypothesis for gene $i$ if
$p_i > 0.05$.

If the null hypothesis is true (i.e. the genes are not differentially expressed) then 
$p_i \leq p_{max}$ in 5% of the cases
```{r hist.p.val.ab, fig.height=4, echo=FALSE}
hist(p.val, nclass=20)
abline(h=length(p.val)/20, col="red")
```

## Controlling False Discovery Rate
To avoid this problem, *p*-values are adjusted (see Wikipedia)
```{r qval, cache=TRUE}
q.val <- p.adjust(p.val, method = "fdr")
```
```{r plot.q, echo=FALSE, cache=TRUE}
plot(p.val, q.val, pch=16, cex=0.4); abline(0,1,col="red")
```

## Adjusted Volcano Plot
```{r volcano.q, cache=TRUE, echo=FALSE, fig.height=4}
plot(lfc, -log10(q.val), pch=16, cex=0.4)
```

+ Biological significance: $LFC_i$
+ Statistical significance: $q_i$

## Adjusted Volcano Plot with thresholds
```{r volcano.q2, cache=TRUE, echo=FALSE, fig.height=4}
plot(lfc, -log10(q.val), pch=16, cex=0.4)
abline(h=-log10(0.05), col="red")
abline(v=1, col="red")
abline(v=-1, col="red")
```

+ Biological significance: $LFC_i>1$ or $LFC_i< -1$
+ Statistical significance: $q_i<0.05$

