---
title: "Installing extra R packages"
author: "Andres Aravena"
date: "14/05/2015"
output: html_document
---

R packages are sets of functions that extend the basic capabilities of R.

They are developed by different people with different motivations. Sometimes two packages provide the same functionality, sometimes they have different functions with the same name.

R comes initially with a set of packages, that you can see opening the **Packages** tab in the lower right quadrant of *Rstudio*. To load them in your session, and make the package functions available to your program, you use the command
```
library(PACKAGE_NAME)
```
using the corresponding `PACKAGE_NAME`. For example, the package *RColorBrewer* has function to create color palettes to produce nice graphs, but its functions are not automatically available in the R session. For example
```{r fail,error=TRUE}
display.brewer.all()
```
To make this function available we have to do
```{r, fig.height=6}
library(RColorBrewer)
display.brewer.all()
```
In summary, the command `library()` loads an external package into the current R session.

## Installing new packages
What if the package is not available in the computer?

R can only load packages that have been previously *installed* in the hard disk.